Efficient ancestry and mutation simulation with msprime 1.0

### 2021-08-13 paper.tex
last few sections

\subsection*{Demography}
% TODO should we put in some citations here for the models? Hardly seems
% necessary.
One of the key applications of population genetic simulations is to generate
data for complex demographies. Beyond idealised cases such as stepping-stone or
island models, or specialised cases such as isolation-with-migration models,
analytical results are rarely possible. Simulation is therefore integral to the
development and evaluation of methods for demographic inference. The demography
model in \msprime\ is directly derived from the approach used in \ms, and
supports an arbitrary number of randomly mating populations exchanging migrants
at specified rates. A range of demographic events are supported, which allow for
varying population sizes and growth rates, changing migration rates over time,
as well as population splits, admixtures and pulse migrations.
The location of sampled lineages
can be tracked through time in as much detail as required: each tree node is
automatically associated with the population in which it arose, the location of
lineages can be recorded at any given time via census events, or every lineage
migration can be recorded.
Large demographic models can be simulated efficiently in version \msprime\ 1.0,
since we only consider populations that contain lineages and
have non-zero migration rates when generating migration event
waiting times. This is a considerable improvement
over version 0.x, which scaled quadratically with the number of populations.

@@

We have implemented instantaneous bottlenecks in \msprime~1.0
using a variant of Hudson's linear time single-locus coalescent
algorithm~\citep{hudson1990gene}. Instantaneous bottlenecks are specified
by adding events to the Demography object (see the Demography section)
and can be used in combination with any other demographic modelling
features. We have validated the results of these simulations by comparing
against analytical expectations for coalescence times and the
site frequency spectrum~\citep{bunnefeld2015inferring}.

@@

Kingman's coalescent assumes that only two ancestral lineages can merge at
each merger event. Although this is generally a reasonable approximation, there
are certain situations in which the underlying mathematical assumptions are
violated. For example in certain highly fecund organisms
\citep{hedgecock_94,B94,HP11,A04,irwin16}, where individuals have the ability
to produce numbers of offspring on the order of the population size and
therefore a few individuals may produce the bulk of the offspring in any given
generation~\citep{hedgecock_94}. These population dynamics violate basic
assumptions of the Kingman coalescent, and are better modelled by
`multiple-merger' coalescents \citep{DK99,P99,S99,S00,MS01}, in which more than
two lineages can merge in a given event. Multiple-merger coalescent processes
have also been shown to be relevant for modelling the effects of selection on
gene genealogies \citep{Gillespie909,DS04}.

Although multiple merger coalescents have been of significant theoretical
interest for around two decades, there has been little practical software
available to simulate these models.
\cite{kelleher2013coalescent,kelleher2014coalescent} developed packages to
simulate a related spatial continuum model~\citep{barton2010new},
\cite{zhu2015hybrid} simulate genealogies within a species tree
based on a multiple-merger model, and
\cite{becheler2020occupancy} provide a general method for simulating
multiple merger processes
as part of the Quetzal framework~\citep{becheler2019quetzal}.
The \texttt{Beta-Xi-Sim} simulator~\citep{koskela2018multi,koskela2019robust}
includes a number of extensions to the Beta-coalescent.
None of these methods work with large genomes, and very little work
has been performed on simulating multiple merger processes with recombination.

We have added two multiple merger coalescent models in \msprime\ 1.0, the
Beta-coalescent and ``Dirac''-coalescent, allowing us to simulate
such models with recombination efficiently for the first time.
These simulation models have been extensively validated against
analytical results from the site frequency
spectrum~\citep{birkner2013statistical,blath2016site,hobolth2019phase}
 as well as more general properties of coalescent processes.
Please see the Appendix for more details and model derivations.

@@

% Axis titles
\node (topLabel) at ($(topLeft) + 5*(genUnit) + (0,-1.3)$) {$\scriptsize\textrm{Genome position}$};
\end{tikzpicture}
\end{subfigure}
\caption{\label{fig-arg} (A) A simple ARG in which a recombination
occurs at position 0.3; (B) the equivalent topology depicted as a tree
sequence, including the recombination node; (C) the same tree sequence
topology ``simplified'' down to its minimal tree sequence representation.}
\end{figure}

For our purposes, an ARG is a realisation of the coalescent with
recombination, in the Griffiths (little ARG) sense.
As described in detail by~\cite{kelleher2016efficient}, Hudson's algorithm
works by dynamically generating and traversing a little ARG.
The graph is not explicitly represented in memory, but is partially
present through the extant lineages and the ancestral material they carry
over time. We do not output the graph directly, but
rather store the information required to recover the genealogical
history as nodes and edges in a tree sequence.
This is far more efficient than outputting the simulated ARG in its entirety.
For a given scaled recombination rate $\rho$
(setting aside the dependency on the sample size $n$)
we know from Eq.~\eqref{eqn-hsw-bound} that the number of nodes
in an ARG is $O(\rho^2)$,
whereas the size of the tree sequence encoding is
$O(\rho)$~\citep{kelleher2016efficient}.
This difference between a
quadratic and a linear dependency on $\rho$ is profound, and shows why
large simulations cannot output an ARG in practice.

@@

Following the approach taken in \discoal, we
have implemented selective sweeps in \msprime\ by generating
sweep trajectories from a jump process approximation to the conditional diffusion
of an allele bound for
fixation~\citep{coop2004ancestral}---see the appendix for more details.
This implementation facilitates future changes to generalise sweep models to
changing population sizes as well as selection from standing variation,
both of which are features that are planned for the near future.
Moreover because this implements the basic infrastructure for the structured
coalescent, adding support for similar situations,
such as inversions~\citep{peischl2013sequential}, is straightforward.
Fig.~\ref{fig-selection-perf} shows that \msprime\ has far better
CPU time and memory performance than \discoal. We have previously shown
that \msprime\ is much more efficient than \msms\ and \cosiTwo\
when simulating without selection~\citep{kelleher2016efficient},
thus we believe \msprime\ to be the most efficient
coalescent simulator when simulating sweeps.

\subsection*{Discrete time Wright-Fisher}
The coalescent is an idealised model and makes many simplifying assumptions,
but it is often surprisingly robust to violations of these
assumptions~\citep{wakeley2012gene}. One situation in which the
model does break down is the combination of large sample size
and long recombining genomes, where the large
number of recombination events in the recent past results in
more than the biologically possible $2^t$ ancestors in
$t$ diploid generations~\citep{nelson2020accounting}.
This pathological behaviour results in
Identity-by-descent,
long-range linkage disequilibrium (LD), and ancestry patterns deviating from
Wright-Fisher expectations, and the bias grows with larger sample
sizes~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting}.
Precisely this problem occurs when simulating modern human datasets,
and we have implemented a Discrete Time Wright-Fisher (DTWF) model
in \msprime\ to address the issue. The DTWF simulates backwards in
time generation-by-generation so that each gamete has a unique
diploid parent, and multiple recombinations within a generation results in
crossover events between the same two parental haploid copies.
The method is described in more detail by~\cite{nelson2020accounting}.

@@

\subsection*{Integration with forward simulators}
A unique feature of \msprime\ is its ability to simulate
genetic ancestries by extending an existing partial
genetic ancestry. Given a tree sequence that
is complete up until time $t$ as input
(where marginal trees may or may not have fully coalesced),
\msprime\ can efficiently obtain the segments of ancestral material
present at this time, and then run the simulation backwards in time from there.
This allows a simulated ancestry to be produced by any
number of different processes across disjoint time slices.
In practice this feature is used to ``complete''
forwards-time ancestry simulations~\citep{kelleher2018efficient}
that may have not fully coalesced. This process
(``recapitation'') can be orders of magnitude faster than
the standard approach of neutral burn-in; see
\cite{haller2018tree} for more details and examples.
This interoperability between simulators, where a partial ancestry
simulation produced by \SLiM~\citep{haller2019slim}
or \fwdpy~\citep{thornton2014cpp} can be picked up and completed
by another simulator, with complete information
retained---at scale---is unprecedented.
This interoperability is indicating an opportunity for
other forward genetic simulators ~\citep[e.g.][]{Gaynor2021AlphaSimR}
to leverage tree sequence data format and associated tools.

\subsection*{Development model}
\texttt{Msprime} has a large number of features, encompassing
the functionality of several more specialised simulators
while maintaining excellent performance.
It is developed by a geographically distributed team of volunteers under an
open source community development model, with a strong emphasis
on code quality, correctness, good documentation, and inclusive development.
As in any large code base,
unit tests play a key role in ensuring that new additions behave
as expected and \msprime\ has an extensive suite.
As of the 1.0.0 release \msprime\ consists of around 13K lines of C and 11K
lines of Python, with suites of 122 C tests (7K lines of code) and
1350 Python tests (22K lines of code). These tests are run
automatically on different operating systems on each pull request
(where a contributor proposes a code change), using standard Continuous
Integration (CI) methodology. Other CI services check code formatting
and for common errors, as well as producing reports on the level of
test coverage, e.g., pinpointing areas where additional unit tests may
be required.

@@

% We can simulate so much more now

This release makes \msprime\ able to produce simulations
with a great deal more biological realism, without a loss in speed and efficiency.
This ability will allow researchers to perform more robust
power analyses, more reliably test new methods,
and carry out more reliable inferences,
without worrying that gene conversion or back mutation might be sabotaging
their applications to real data.
These advances have only been possible thanks to a distributed,
collaborative model of software development,
and the work of many people.
We hope the tool continues to be useful to many.

@@

% THERE IS A BETTER WAY!!!
Scientific software is vital apparatus, and must be engineered
to a high quality if we are to trust its results. There is a growing
realisation across the sciences \citet[OTHER CITATIONS]{siepel2019challenges}
% Maybe cite: https://www.biorxiv.org/content/10.1101/092205v3
that investing in
shared community infrastructure produces better results than a
proliferation of individually maintained tools, allowing scientists
to focus on their specific questions rather than software engineering.
\texttt{Msprime} 1.0 is the result of such a community process,
with features added by motivated users, taking advantage of the
established development practices and infrastructure.
% I'm going out on a limb here - do folks agree?
% PLR- personally I have learned SO MUCH, and I'm passing it on to others
Software development in a welcoming community,
with mentorship by experienced developers,
is a hugely useful experience for many users.
The skills that contributors learn
can lead to greatly increased productivity in subsequent
work (e.g., through more reliable code and better debugging skills).
We hope that users who find that
features they require are missing will continue to contribute to
\msprime, leading to a community project that is both high quality
and sustainable in the long term.

>>>>

\subsection*{Demography}
\label{sec-demography}
% TODO should we put in some citations here for the models? Hardly seems
% necessary.
One of the key applications of population genetic simulations is to generate
data for complex demographies. Beyond idealised cases such as stepping-stone or
island models, or specialised cases such as isolation-with-migration models,
analytical results are rarely possible. Simulation is therefore integral to the
development and evaluation of methods for demographic inference. The demography
model in \msprime\ is directly derived from the approach used in \ms, and
supports an arbitrary number of randomly mating populations exchanging migrants
at specified rates. A range of demographic events are supported, which allow for
varying population sizes and growth rates, changing migration rates over time,
as well as population splits, admixtures and pulse migrations.
The location of sampled lineages
can be tracked through time in as much detail as required: each tree node is
automatically associated with the population in which it arose, the location of
lineages can be recorded at any given time via census events, or every lineage
migration can be recorded.
Large demographic models can be simulated efficiently in version \msprime\ 1.0,
since we only consider populations that contain lineages and
have non-zero migration rates when generating migration event
waiting times. This is a considerable improvement
over version 0.x, which scaled quadratically with the number of populations.

@@

We have implemented instantaneous bottlenecks in \msprime~1.0
using a variant of Hudson's linear time single-locus coalescent
algorithm~\citep{hudson1990gene}. Instantaneous bottlenecks are specified
by adding events to the \texttt{Demography} class
(see the \nameref{sec-demography} section)
and can be used in combination with any other demographic modelling
features. We have validated the results of these simulations by comparing
against analytical expectations for coalescence times and the
site frequency spectrum~\citep{bunnefeld2015inferring}.

@@

Kingman's coalescent assumes that only two ancestral lineages can merge at
each merger event. Although this is generally a reasonable approximation, there
are certain situations in which the underlying mathematical assumptions are
violated. For example in certain highly fecund organisms
\citep{hedgecock_94,B94,HP11,A04,irwin16}, where individuals have the ability
to produce numbers of offspring on the order of the population size and
therefore a few individuals may produce the bulk of the offspring in any given
generation~\citep{hedgecock_94}. These population dynamics violate basic
assumptions of the Kingman coalescent, and are better modelled by
`multiple-merger' coalescents \citep{DK99,P99,S99,S00,MS01}, in which more than
two lineages can merge in a given event. Multiple-merger coalescent processes
have also been shown to be relevant for modelling the effects of selection on
gene genealogies
\citep{Gillespie909,DS04,desai2013genetic,neher2013genealogies,schweinsberg2017rigorous}.

Although multiple merger coalescents have been of significant theoretical
interest for around two decades, there has been little practical software
available to simulate these models.
\cite{kelleher2013coalescent,kelleher2014coalescent} developed packages to
simulate a related spatial continuum model~\citep{barton2010new},
\cite{zhu2015hybrid} simulate genealogies within a species tree
based on a multiple-merger model, and
\cite{becheler2020occupancy} provide a general method for simulating
multiple merger processes
as part of the Quetzal framework~\citep{becheler2019quetzal}.
The \texttt{Beta-Xi-Sim} simulator~\citep{koskela2018multi,koskela2019robust}
also includes a number of extensions to the Beta-coalescent.
None of these methods work with large genomes, and very little work
has been performed on simulating multiple merger processes with recombination.

We have added two multiple merger coalescent models in \msprime\ 1.0, the
Beta-coalescent and ``Dirac''-coalescent, allowing us to simulate
such models with recombination efficiently for the first time.
These simulation models have been extensively validated against
analytical results from the site frequency
spectrum~\citep{birkner2013statistical,blath2016site,hobolth2019phase}
 as well as more general properties of coalescent processes.
See the Appendix for more details and model derivations.

@@

% Axis titles
\node (topLabel) at ($(topLeft) + 5*(genUnit) + (0,-1.3)$) {$\scriptsize\textrm{Genome position}$};
\end{tikzpicture}
\end{subfigure}
\caption{\label{fig-arg} (A) A simple ARG in which a recombination
occurs at position 0.3; (B) the equivalent topology depicted as a tree
sequence, including the recombination node; (C) the same tree sequence
topology ``simplified'' down to its minimal tree sequence representation.
Note that the internal nodes have been renumbered in the simplified
representation, so that, e.g., node 5 in (C) corresponds to node 6 in
(A) and (B).}
\end{figure}

For our purposes, an ARG is a realisation of the coalescent with
recombination, in the Griffiths (little ARG) sense.
As described in detail by~\cite{kelleher2016efficient}, Hudson's algorithm
works by dynamically traversing a little ARG.
The graph is not explicitly represented in memory, but is partially
present through the extant lineages and the ancestral material they carry
over time. We do not output the graph directly, but
rather store the information required to recover the genealogical
history as nodes and edges in a tree sequence.
This is far more efficient than outputting the simulated ARG in its entirety.
For a given scaled recombination rate $\rho$
(setting aside the dependency on the sample size $n$)
we know from Eq.~\eqref{eqn-hsw-bound} that the number of nodes
in an ARG is $O(\rho^2)$,
whereas the size of the tree sequence encoding is
$O(\rho)$~\citep{kelleher2016efficient}.
This difference between a
quadratic and a linear dependency on $\rho$ is profound, and shows why
large simulations cannot output an ARG in practice.

@@

Following the approach taken in \discoal, we
have implemented selective sweeps in \msprime\ by generating
sweep trajectories from a jump process approximation to the conditional diffusion
of an allele bound for
fixation~\citep{coop2004ancestral}---see the Appendix for more details.
This implementation facilitates future changes to generalise sweep models to
changing population sizes as well as selection from standing variation.
Moreover because this implements the basic infrastructure for the structured
coalescent, adding support for similar situations,
such as inversions~\citep{peischl2013sequential}, is straightforward.
Fig.~\ref{fig-selection-perf} shows that \msprime\ has far better
CPU time and memory performance than \discoal. We have previously shown
that \msprime\ is much more efficient than \msms\ and \cosiTwo\
when simulating without selection~\citep{kelleher2016efficient}.
% jk - this just seems clumsy to me: surely the conclusion is clear?
% and we therefore believe \msprime\ to be the most efficient
% coalescent simulator when simulating sweeps.

\subsection*{Discrete time Wright-Fisher}
The coalescent is an idealised model and makes many simplifying assumptions,
but it is often surprisingly robust to violations of these
assumptions~\citep{wakeley2012gene}. One situation in which the
model does break down is the combination of large sample size
and long recombining genomes, where the large
number of recombination events in the recent past results in
more than the biologically possible $2^t$ ancestors in
$t$ diploid generations~\citep{nelson2020accounting}.
This pathological behaviour results in
identity-by-descent,
long-range linkage disequilibrium and ancestry patterns deviating from
Wright-Fisher expectations, and the bias grows with larger sample
sizes~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting}.
Precisely this problem occurs when simulating modern human datasets,
and we have implemented a Discrete Time Wright-Fisher (DTWF) model
in \msprime\ to address the issue. The DTWF simulates backwards in
time generation-by-generation so that each gamete has a unique
diploid parent, and multiple recombinations within a generation results in
crossover events between the same two parental haploid copies.
The method is described in more detail by~\cite{nelson2020accounting}.

@@

\subsection*{Integration with forward simulators}
A unique feature of \msprime\ is its ability to simulate
genetic ancestries by extending an existing partial
genetic ancestry. Given a tree sequence that
is complete up until time $t$ ago as input
(where marginal trees may or may not have fully coalesced),
\msprime\ can efficiently obtain the segments of ancestral material
present at this time, and then run the simulation backwards in time from there.
This allows a simulated ancestry to be produced by any
number of different processes across disjoint time slices.
In practice this feature is used to ``complete''
forwards-time ancestry simulations~\citep{kelleher2018efficient}
that may have not fully coalesced. This process
(``recapitation'') can be orders of magnitude faster than
the standard approach of neutral burn-in; see
\cite{haller2018tree} for more details and examples.
This interoperability between simulators, where a partial ancestry
simulation produced by \SLiM~\citep{haller2019slim}
or \fwdpy~\citep{thornton2014cpp} can be picked up and completed
by another simulator, with complete information
retained---at scale---is unprecedented.
This interoperability indicates an opportunity for
other forward genetic simulators ~\citep[e.g.][]{Gaynor2021AlphaSimR}
to leverage the tree sequence data format and associated tools.

\subsection*{Development model}
\texttt{Msprime} has a large number of features, encompassing
the functionality of several more specialised simulators
while maintaining excellent performance.
It is developed by a geographically distributed team of volunteers under an
open source community development model, with a strong emphasis
on code quality, correctness, good documentation, and inclusive development.
As in any large code base,
unit tests play a key role in ensuring that new additions behave
as expected and \msprime\ has an extensive suite.
As of the 1.0.0 release \msprime\ consists of around 13K lines of C and 11K
lines of Python, with suites of 122 C tests (7K lines of code) and
1350 Python tests (22K lines of code). These tests are run
automatically on different operating systems on each pull request
(where a contributor proposes a code change), using standard Continuous
Integration (CI) methodology. Other CI services
check for common errors, code formatting issues, and produce reports on
the level of test coverage for the proposed change.

@@

% We can simulate so much more now
This release makes \msprime\ able to produce simulations
with a great deal more biological realism, without a loss in speed and efficiency.
This ability will allow researchers to perform more robust
power analyses, more reliably test new methods,
and carry out more reliable inferences,
without worrying that gene conversion or back mutation might be sabotaging
their applications to real data.
These advances have only been possible thanks to a distributed,
collaborative model of software development,
and the work of many people.
We hope the tool continues to be useful to many.

@@

% THERE IS A BETTER WAY!!!
Scientific software is vital apparatus, and must be engineered
to a high quality if we are to trust its results. There is a growing
realisation across the sciences
% More citations?
\citep{siepel2019challenges,harris2020array,gardner2021sustained}
that investing in
shared community infrastructure produces better results than a
proliferation of individually maintained tools, allowing scientists
to focus on their specific questions rather than software engineering.
\texttt{Msprime} 1.0 is the result of such a community process,
with features added by motivated users, taking advantage of the
established development practices and infrastructure.
% JK: I'm going out on a limb here - do folks agree?
% PLR- personally I have learned SO MUCH, and I'm passing it on to others
Software development in a welcoming community,
with mentorship by experienced developers,
is a useful experience for many users.
The skills that contributors learn
can lead to greatly increased productivity in subsequent
work (e.g., through more reliable code and better debugging skills).
We hope that users who find that
features they require are missing will continue to contribute to
\msprime, leading to a community project that is both high quality
and sustainable in the long term.
